Discussion Draft

### 2020-03-31 doc/Discussion.rmd
small updates

To synthesize our results, we decded to systematically compare models which have been proposed for explaining the evolution of heterochiasmy or meiotic machinery. There are two main types of models: modifers (indirect) modifers or functional (cell physiology based models). We attempted to test the hypotheses (mental test for if the model) (Table X).

**the models**  We examined 3 modifer models; haploid selection, two locus model, and SACE. We also examined and propose cell physiology based models: the Chromosome Ocsilation (COM) model and a model focused on the spindle differences between males and females.

<we left out Beam film model for XX reasons>

<heterochiasmy refers to different numbers of chiasmata - and recently has been extended to number of crossovers -- this view / charaecterization is slightly incomplete, since it considers just one cell stage and 1 value (a count))
We argue that the sex differences (broaden the definition of heterochiasmy or use a different word) -- which includes more aspects that show conserved sexes differences>

<Our results indicate, sexes have distinct evolutionary trajectories for genome wide recombinaiotn rate (and landscape) in males nd females).

Conserved features of the landscape may play a role in influencing/keeping the sex specific trajectories. In our examination of the panel of mouse strains, we note several features of the recobmination landscape with conserved sexual dirmophism pattern. 
That is the sex within sex these features show no evolution and consistant pattern of variation.>

We focus first on the results for sexually dimorphic traits.

#reversed hetC direction

The common direction of heterochiasmy is female biased, however male biased heterochiasmy has significant number of observations, seen in cattle, marsupial, and (more to the list). Our results are novel in the sense that the evolutionary scale of the male-specific is short.

The prediction from two of the modifier models, haploid selection and SACE are that males will generally evolve lower recombination rates overall compared to females. <cases with exceptions.? we don't have evidence or reason to predict these exception cases are held>
The two locus modifier model  does parameterize cases for male biased heterochiasmy, dependant on the stage of action for the driver (MI or MII) and the linkage status (see table 1 Brandvain and Coop).

There are no prediction for male biased heterochiasmy evolution direction under the COM <pairing model>, but it should be noted that this model wasn't designed to incorporate this type of data. This model were design to explain the positive interference pattern for of the recombination landscape by incorporating a cell mechanics of rapid prophase movements (RPM) that are thought to facilitate homolog pairing in many species.

@@

(Lane and Kauppi) [@laneKauppi2019]

@@

centromeres have sinlge and opposite pole connection (synthellic?) and that the sister cohesion connecting the homologs -- sets up the tension -- opposes the pull from each of the poles.

Efficient normal SAC -- works by pausing entry into anaphase or triggers apoptosis -- if a unpaired / tension-lacking tetrad is detected
(sperm SAC is finer tuned (can detect 1-2 tetrads lacking tension -- ), while the female SAC  (has a higher treshold for being triggered  (4-5?)
(Land Kauppi,

@@

# 'Typical landscape'

#-long observed
#measured from a variety of approaches,
#larger variety of species

#-Exception to landscape (marisupials, birds?)

#which models make clear predctions for typical landscape?
#SACE
#drivers
#spindle differences

Our results on the broad scale pattern of crossover placement are generally supported by the literature. As reviewed in [@sardell_sex_2020], sexual dirmorohism in the broad scale landscape is a highly conserved trait.  However there are exceptions, (marsupial, bird?)
the popgen models
The two locus modifier model and SACE model make predictions for (sexual dimorphism ) in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks respectively.  
functional models
The two mechanical models also both predict sexual dimorphism in the recombination landscape for distinct reasons. The chromosome pairing based model (COM), predicts the sex differences in the positioning of crossovers is due to a combination of the length of the axis and intensity of the RPMs.  The spindle based model proposed that sexual dimorphism in the recombination landscape is driven by difference strength in the spindle assembly checkpoint (Lane Kauppi). The relaxed selection imposed by an ineffective SAC results in a uniform distribution of positions (while maintaining positive interference). Whereas,  in males, the directional selection could cause decreased variance in the recombination landscape via telomeric positioning of crossovers. Under the telomeric bias  recombination landscape, entry into anaphase may be more synchoriznes (and faster) given that the reductional separation of homologs is dependent on a minimal amount of sister cohesion connecting the homologs (Figure X).

[@sardell_sex_2020].

@@

#our results, female more (greater than tech error) male no significant difference
#between cell variation

#between cell variance (Veller Kleckner), mechnism
#evolution / popgen model from (veller)

#models that predict more within animal variance
#none of the models expliciltly predict fr traits at the between cell level
#so any between cell variation -- patterns is due to DIFFERENCES IN THE STRENGTH OF SELECTION

## connection on evolution of transcriptional noise
# transcriptional noise and complex traits -- both show the same repsonses to different modes of selection

@@

@@

```{r chromatin.organization, include=FALSE}
#1. basics and expectations for chromatin organization
#seen in humans, mice (signal comes through even with cell to cell variation in chromatin compaction / organization)

#2. how the model treat this limiting factor

#3. consequence for interference (general patterns) is that interference measured as (mb) will be stronger if DNA loops are stronger
#prediction for general -- pattern
```

@@

**chromatin organization**

The results of sexual dimorphism in chromatin structure is supported by the literature, especially for mammals [@lynn2002], [@gruhn2013]. 
models
Very few models make predictions for how chromatin organization relates to variation in recombination rates / the recombination landscape.  (the chromatin organization is how chromosomes with the same number of base pairs) are transformed into bottle brush structures with a central axis and most of the genetic material arranged as loops emanating outwards (figure).
Only the COM model states that chromatin structure will be longer in oocytes due to greater cell volume.  This prediction / model might fit broader pattern (such as in At where pollen is the larger cell and has longer axis length) (cahoon libuda) [@cahoonLibuda2019].  
Predictions for this trait are  inconclusive (due to lack of cytology measures from both sexes. We argue that the chromatin structure (and axis length) should be considered as an integral parameter when studying recombination landscape.  The sex differences in chromatin strucutre are closely connected to sex differences  in the interference strength and our results mirror those in Petkov et al.  (prediction, shorter-loop will have stronger interference).

According to the tethered loop model of DSB formation (cite) -- the 'real estate' for COs are the DNA loops - the density of whch is conserved in the axis and loop formation (later the SC str), so the number of loops is equvilent to the length of the axis. **The chromosomes are converted to a 2D space (the axis length of a given chromosome is the limited factor for total CO number)**

2. Only one of the models we examined aligns with our observed results for sexual dimorphism for axis length.

The modifier models do not parameterize this since they do not models / are not built to consider single meiotic events.

**COM** predicts that axis length will be larger in oocytes due to the larger volume -- this would help in the bouquet stage -- may be connected to the 'the search' during homolog pairing.

the **spindle based selection** model does not specific predictions about this aspect -- assumes that the 2D areas for chromosomes are proportional across stages (during the decondensing stage between pachtyene and dipl? / metaphase) -- the chromosomes condenses uniformly across the length.

>>>>

We performed hypothesis testing of models from the literature and adapted here fit the main results from this study. We choose to examine three modifier models and two functional and cell physiology models (table X). It is a bit challenging to test all of these models together since they were built with difference parameters and built to describe different patterns of recombination variation. The modifier models were built to explain heterochiasmy variation between sexes and the COM functional model was built to describe the interference pattern and difference in the recombination landscape between the sexes. In cases where the models can't be extended to predicting results, we list as 'NA'.

# Reversed heterochiasmy direction

```{r reversed.HetC, echo=FALSE}

The common direction of heterochiasmy is female biased, however male biased heterochiasmy has significant number of observations (list). Our results are novel in the sense that the evolutionary scale of the male-specific is short.

The prediction from two of the modifier models, haploid selection and SACE are that males will generally evolve lower recombination rates overall compared to females. <cases with exceptions.? we don't have evidence or reason to predict these exception cases are held>
The two locus modifier model  does parameterize cases for male biased heterochiasmy, dependent on the stage of action for the driver (MI or MII) and the linkage status (see table 1 Brandvain and Coop).

There are no prediction for male biased heterochiasmy evolution direction under the COM <pairing model>, but it should be noted that this model was not't designed to incorporate this type of data. This model were design to explain the positive interference pattern for of the recombination landscape by incorporating a cell mechanics of rapid prophase movements (RPM) that are thought to facilitate homolog pairing in many species.

@@

[@laneKauppi2019]

@@

centromeres have single and opposite pole connection (synthellic?) and that the sister cohesion connecting the homologs -- sets up the tension -- opposes the pull from each of the poles.

Efficient normal SAC -- works by pausing entry into anaphase or triggers apoptosis -- if a unpaired / tension-lacking tetrad is detected
(sperm SAC is finer tuned (can detect 1-2 tetrads lacking tension -- ), while the female SAC  (has a higher threshold for being triggered  (4-5?)
(Land Kauppi,

@@

@@

@@

# Typical landscape

# typical landscape is conserved -- sexually dimorphic meiotic trait (Sardell)

# models

#<spindle model -- the telomeric placement of crossovers, minimizes the amount of sister cohesion connecting homologs>

Our results on the broad scale pattern of crossover placement are generally supported by the literature. As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale landscape is a highly conserved trait.

The two locus modifier model and SACE model make predictions for sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks respectively.

The two mechanical models also both predict sexual dimorphism in the recombination landscape for distinct reasons. The COM model (chromosome pairing based model), predicts the sex differences in the positioning of crossovers is due to a combination of the length of the axis and intensity of the rapid prophase movements (RPMs) (hulten, Rubin et al 2020).

The spindle based model proposed that sexual dimorphism in the recombination landscape is driven by difference strength in the spindle assembly checkpoint (Lane Kauppi). The relaxed selection imposed by an ineffective SAC results in a uniform distribution of positions (while maintaining positive interference).

Just as the difference in sac (checkpoint) strength can lead to disinct evolutionary trajectories of the genome wide recombiation rate for oocytes and spermatocytes, the other side of the same coin is that there is increased variance between cells for crossover number. (weaker / leaky checkpoint )

The telomeric placement of crossovers in the male recombiation landscape, minimizes the amount of sister cohesion connecting homologs. This conserved feature of the male recombination landscape, might act to synchornize the seperation homologs after alignment at metaphase, (and in the degargation of the sister cohesion connecting the homologs. Another conserved feature of gametogenesis is that the timing of spermatocyte metapphase is much shorter of time compared to oocytes.

Whereas, in males, the directional selection could cause decreased variance in the recombination landscape via telomeric positioning of crossovers. Under the telomeric bias  recombination landscape, entry into anaphase may be more synchronizes (and faster) given that the reductional separation of homologs is dependent on a minimal amount of sister cohesion connecting the homologs (Figure X).

@@

```{r chromatin.organization, include=FALSE}

#supporting results from lit
#chromatin organization

#models
# none of the models make predictions for the axis length (chromatin organization)
# only COM predicts that sexual dimorphism in axis length is due to cell vol

@@

The results of sexual dimorphism in chromatin structure is supported by the literature, especially for mammals [@lynn2002], [@gruhn2013].

Very few models make predictions for how chromatin organization relates to variation in recombination rates / the recombination landscape.  (the chromatin organization is how chromosomes with the same number of base pairs) are transformed into bottle brush structures with a central axis and most of the genetic material arranged as loops emanating outwards (figure). 
Only the COM model states that chromatin structure will be longer in oocytes due to greater cell volume.  This prediction / model might fit broader pattern (such as in At where pollen is the larger cell and has longer axis length) [@cahoonLibuda2019].  
Predictions for this trait are  inconclusive (due to lack of cytology measures from both sexes. We argue that the chromatin structure (and axis length) should be considered as an integral parameter when studying recombination landscape.  The sex differences in chromatin strucutre are closely connected to sex differences  in the interference strength and our results mirror those in Petkov et al.  (prediction, shorter-loop will have stronger interference).

According to the tethered loop model of DSB formation (cite) -- the 'real estate' for COs are the DNA loops - the density of whch is conserved in the axis and loop formation (later the SC str), so the number of loops is equvilent to the length of the axis. **The chromosomes are converted to a 2D space (the axis length of a given chromosome is the limited factor for total CO number)**.  Only one of the models COM aligns with our observed results for sexual dimorphism for axis length. **COM** predicts that axis length will be larger in oocytes due to the larger volume.
